---
title: "Eric Jensen Markdown"
author: "Eric Jensen"
date: "February 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Unpacking data and libraries
```{r}
library(ggplot2)
library(dplyr)
library(ggmap)

temp <- tempfile()
download.file("https://s3.amazonaws.com/tripdata/201601-citibike-tripdata.zip", temp)
M1 <- read.csv(unz(temp, "201601-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201602-citibike-tripdata.zip", temp)
M2 <- read.csv(unz(temp, "201602-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201603-citibike-tripdata.zip", temp)
M3 <- read.csv(unz(temp, "201603-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201604-citibike-tripdata.zip", temp)
M4 <- read.csv(unz(temp, "201604-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201605-citibike-tripdata.zip", temp)
M5 <- read.csv(unz(temp, "201605-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201606-citibike-tripdata.zip", temp)
M6 <- read.csv(unz(temp, "201606-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201607-citibike-tripdata.zip", temp)
M7 <- read.csv(unz(temp, "201607-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201608-citibike-tripdata.zip", temp)
M8 <- read.csv(unz(temp, "201608-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201609-citibike-tripdata.zip", temp)
M9 <- read.csv(unz(temp, "201609-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201610-citibike-tripdata.zip", temp)
M10 <- read.csv(unz(temp, "201610-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201611-citibike-tripdata.zip", temp)
M11 <- read.csv(unz(temp, "201611-citibike-tripdata.csv"), header = T)
unlink(temp)

download.file("https://s3.amazonaws.com/tripdata/201612-citibike-tripdata.zip", temp)
M12 <- read.csv(unz(temp, "201612-citibike-tripdata.csv"), header = T)
unlink(temp)

#Updating names for M10-M12 to match the headers of the others
names(M10) <- c("tripduration", "starttime", "stoptime", "start.station.id", "start.station.name", "start.station.latitude", "start.station.longitude", "end.station.id", "end.station.name", "end.station.latitude", "end.station.longitude", "bikeid", "usertype", "birth.year", "gender")
names(M11) <- c("tripduration", "starttime", "stoptime", "start.station.id", "start.station.name", "start.station.latitude", "start.station.longitude", "end.station.id", "end.station.name", "end.station.latitude", "end.station.longitude", "bikeid", "usertype", "birth.year", "gender")
names(M12) <- c("tripduration", "starttime", "stoptime", "start.station.id", "start.station.name", "start.station.latitude", "start.station.longitude", "end.station.id", "end.station.name", "end.station.latitude", "end.station.longitude", "bikeid", "usertype", "birth.year", "gender")

#Combine to single data set
citibike2k16 <- rbind(M1, M2, M3, M4, M5, M6, M7, M8, M9, M10, M11, M12)

```

#Most Frequent Start Station
```{r}
by.start <- citibike2k16 %>%
  group_by(start.station.id, start.station.longitude, start.station.latitude) %>% summarize(frequency = n())
by.start <- as.data.frame(by.start)
names(by.start) <- c("station.id", "longitude", "latitude", "frequency")

by.start <- by.start %>%
  select(station.id, longitude, latitude, frequency) %>%
  arrange(desc(frequency))

by.start.1000 <- by.start[1:1000, ]

#get map base
source = "google"
NYCmap <- get_map("New York City",zoom=15,maptype="roadmap",source= source)

#Dot Plot of top 1000 most popular start stations
ggmap(NYCmap) + geom_point(aes(x=longitude,y=latitude, size = frequency, color=frequency), data=by.start.1000) + geom_point(size=5,alpha=0.3) 

#Barplot of top 10 most popular start stations
by.start.10 <- by.start[1:10, ]

by.start.10$station.id

barplot(by.start.10$frequency, main = "Most Frequent Start Stations", xlab = "Station ID", ylab = "Frequency", names.arg = c("519", "435", "497", "426", "402", "490", "285", "284", "151", "459"), col = "blue")
```

As you can see, the most common start station is station 519 which is Pershing Square North

#Most Frequent End Station
```{r}
#end station frequency
by.end <- citibike2k16 %>%
    group_by(end.station.id, end.station.longitude, end.station.latitude) %>% summarise(n=n())
by.end <- as.data.frame(by.end)
names(by.end) <-  c("station.id", "longitude", "latitude", "frequency")

by.end <- by.end %>%
  select(station.id, longitude, latitude, frequency) %>%
  arrange(desc(frequency))

by.end.1000 <- by.end[1:1000, ]

#get map base
source = "google"
NYCmap <- get_map("New York City",zoom=15,maptype="roadmap",source= source)

#Dot Plot of top 1000 most popular start stations
ggmap(NYCmap) + geom_point(aes(x=longitude,y=latitude, size = frequency, color=frequency), data=by.end.1000) + geom_point(size=5,alpha=0.3) 

#Barplot of top 10 most popular start stations
by.end.10 <- by.end[1:10, ]

barplot(by.start.10$frequency, main = "Most Frequent End Stations", xlab = "Station ID", ylab = "Frequency", names.arg = c("519", "402", "426", "497", "435", "285", "490", "284", "459", "477"), col = "blue")
```

Similar to the start station, the most common end station is 519 which is Pershing Square North.

#Most Frequent Station
```{r}

#merging data to find most frequent
by.start.end <- merge(by.start, by.end, by = "station.id")
by.start.end$tot.freq <- by.start.end$frequency.x + by.start.end$frequency.y
by.start.end$longitude.y <- NULL
by.start.end$latitude.y <- NULL
names(by.start.end) <-  c("station.id", "longitude", "latitude", "start_frequency", "end_frequency", "frequency")

#arranging and pulling top 1000 most frequent stations
by.start.end <- by.start.end %>%
  select(station.id, longitude, latitude, frequency) %>%
  arrange(desc(frequency))
by.start.end.1000 <- by.start.end[1:1000, ]

#get map base
source = "google"
NYCmap <- get_map("New York City",zoom=15,maptype="roadmap",source= source)

#Dot Plot of top 1000 most popular stations
ggmap(NYCmap) + geom_point(aes(x=longitude, y=latitude, size = frequency, color=frequency), data=by.start.end.1000) + geom_point(size=5,alpha=0.3) 

#Barplot of most frequently used station
by.start.end.10 <- by.start.end.1000[1:10, ]

barplot(by.start.end.10$frequency, main = "Most Frequented Stations", xlab = "Station ID", ylab = "Frequency", names.arg = c("519", "435", "402", "497", "426",
                                   "285", "490", "284", "151", "459"), col = "blue")
```
As expected, the most common station is 519, Pershing Square North, which was also the most common start and end station

#Most Frequent Routes
```{r}
#merging start and end station
citibike2k16.routes <- citibike2k16
citibike2k16.routes$station.id <- paste(citibike2k16.routes$start.station.id, citibike2k16.routes$end.station.id, sep="_")

#Finding route frequency
by.route <- citibike2k16.routes %>%
  group_by(station.id, start.station.latitude, start.station.longitude, end.station.latitude, end.station.longitude) %>% summarise(frequency = n())
by.route <- as.data.frame(by.route)

#Arranging by route and top 1000 routes
by.route <- by.route %>%
  select(station.id, start.station.longitude, start.station.latitude, end.station.longitude, end.station.latitude, frequency) %>%
  arrange(desc(frequency))
by.route.1000 <- by.route[1:1000, ]

#most common path map
ggmap(NYCmap) +
  geom_polygon(data = by.route.1000,
               aes(x= start.station.longitude, y = start.station.latitude, group = station.id),
               fill = "#080808", color = "#080808", alpha = 0.5) +
  geom_leg(data = by.route.1000,
           aes(x = start.station.longitude, xend = end.station.longitude,
           y = start.station.latitude, yend = end.station.latitude,
           size = frequency, color = frequency, alpha = frequency)) + 
  geom_point(data = by.route.1000,
             aes(x = start.station.longitude, y = start.station.latitude),
             color = "#ffa500", alpha = 0.5, size = 1) +
  scale_size_continuous(range = c(0.3, 3)) + 
  scale_color_gradient(low = "#555555", high = "#ffffff", trans = "sqrt") +
  scale_alpha_continuous(range = c(0.6, 1), trans = "sqrt")

by.route.10 <- by.route.1000[1:10, ]

#Barplot of most frequented routes
barplot(by.route.10$frequency, names.arg = c("2006_2006", "2006_3143", "387_387",   "514_426", "435_509", "519_492", "3150_3147", "519_477", "519_491", "3093_2002"), main = "Most Common Routes", ylab = "Frequency", col = "Blue",las = 2)
```

The most common route is 2006_2006 which is Central Park S & 6th Ave. This makes sense because the most common route is for people to ride their bikes around Central Park.
